An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-23 paper.tex
Clean up definitions leading up to the description of BFS

\begin{defn}
An \textit{evolving graph} $G_n$ is a sequence of
(static) graphs
$G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]} \rangle$ with associated
time labels $t_1, t_2, \ldots, t_n$ respectively.
Each $G^{[t]} = (V^{[t]}, E^{[t]})$ represents a (static) graph labeled by a
time $t$.
\end{defn}

@@

We assume no particular relation between the node and edge sets for each static
graph $G^{[t]} = (V^{[t]}, E^{[t]})$.

\begin{defn}
A \textit{temporal node} is a pair $(v, t)$, where $v\in V^{[t]}$ is a node at
a time $t$.
\end{defn}

\begin{defn}
An \textit{active node} is a temporal node $(v,t)$ if there exists at least one
edge $e\in E^{[t]}$ that connects $v\in V^{[t]}$ to another node $w\in V^{[t]}$,
$w\ne v$.

An \textit{inactive node} is a temporal node that is not an active node.
\end{defn}

@@

\begin{defn}
A \textit{temporal path on an evolving graph $G_n$ from temporal node $(v_1, t_1)$
to temporal node $(v_m, t_m)$} is a time-ordered sequence of temporal notes,
$\langle (v_1, t_1), (v_2, t_2), \ldots, (v_m, t_m) \rangle$.
Here, time ordering means that $t_1 \leq t_2 \leq \cdots \leq t_m$ and
$v_i = v_j$ iff $t_i \ne t_j$.
The length of a temporal path is the length of the sequence.
\end{defn}

This definition of a temporal path is essentially that of the dynamic walk
in \cite{gphe11,grihig13}.

Grindrod et al.'s model  \cite{gphe11} \cite{grihig13} is very similar to ours,
where an evolving graph is considered as a sequence of adjacency matrices.
% We present a preliminary discussion of using the model to mine citation networks.

% why BFS is important?

% Given a graph and a source node $s$,
% the breadth-first search (BFS) systematically discover every node that is reachable from $s$.
% As the name suggested, BFS expands the frontier of discovered regions
% uniformly across the breadth, i.e., the algorithm discovers all nodes
% that are $k$ walks away from $s$ before discovering any nodes that are
% $k+1$ walks away.
A temporal path is the basic element to
 define almost  all important evolving graph algorithms.
Matrix-matrix multiplication between consecutive adjacency matrices
were commonly used as a means of determining temporal paths \cite{gphe11,grihig13}. However, we show
in one specific case it fails to include all the temporal paths and thus should not be used
for searching evolving graphs in general.

Breadth-first search (BFS) is probably the most basic graph searching algorithm and
it is also the building block of other graph algorithms like minimum spanning tree
and shortest path \cite{ckrs09}.
Despite the fact that graph metrics \cite{ntmm13,tmml09,tsmm09} and
various centrality measures like Katz centrality~\cite{gphe11,grihig13},
betweenness \cite{alhi15}, and closeness \cite{tmml10})
have been generalized for evolving graphs,
a BFS for evolving graphs, which we shall call it \emph{evolving-graph BFS},
was never explicitly described or verified.
Therefore, there is a need for clear presentations of the algorithm.

In this paper,  we present descriptions of an evolving-graph BFS algorithm, but its main concern is with its sparse matrix
representation and linear algebraic computations.
We prove that the two algorithms (graph notation and algebra notation) are equivalent.

% GraphBlas related ideas

% This paper is organized as follows. The evolving-graph BFS algorithm is described
% and analyzed using graph notation in Section \ref{sec:breadth-first-search}. Section \ref{sec:evolv-graph-algor} considers the algebraic representation of the algorithm:
% A sparse matrix representation
% of an evolving graph is presented in \ref{sec:representation}.
% The algorithm is presented in \ref{sec:breadth-first-search-1} and
% the properties of the search tree is discussed in \ref{sec:short-temp-path}.
% The connection between a temporal path and a matrix-matrix multiplication is considered and discussed in \ref{sec:matr-matr-mult}. Finally,
% an application of the evolving-graph BFS in mining citation networks is described in Section \ref{sec:applications}.

We would like a BFS on evolving graphs to work similar to that on graphs: the search starts with some source node $(s,t)$, called \emph{root}. It constructs a tree through discovering all nodes that are reachable by
temporal paths of length $k$ before moving to all nodes that are reachable by
 temporal paths of length $k+1$.
By construction, the time stamp of an active node is always not later than
that of its descendant. Suppose we
start with node $(1, t_1)$ in Figure \ref{fig:eg_shortest_path}.
After $1$ temporal path from $(1, t_1)$, we can reach
$(2, t_1)$ and $(1, t_2)$. After $2$ temporal paths, we can reach
$(2, t_1), (1, t_2), (2, t_2)$ and $(3, t_2)$.
% This example motivates us to define the reachable set of
% a node-timestamp pair.
Formally, we define the \emph{forward-neighbors} of a node $(v, t)$ to
be the set of all the reachable nodes after $1$ temporal path from
$(v, t)$, i.e., the nodes pointed by $v$ at time stamp $t$ and
the active node $v$ itself at later time stamps. In Figure
\ref{fig:eg_shortest_path},
the forward-neighbors of $(1, t_1)$  are $(2, t_1)$ and $(1, t_2)$ and
the forward-neighbors of $(2, t_1)$ is $(2, t_3)$.

Given an evolving graph $G_n$ and a starting active node $(s,t)$, Algorithm \ref{alg:bfs}
returns all the reachable active nodes and their distances from $(s,t)$.
For Figure \ref{fig:eg_shortest_path}, suppose we start with $(1, t_2)$, then
$frontier$ is $\langle(1,t_2)\rangle$ after step $5$.  Since the only forward-neighbor of
$(1,t_2)$ is $(3,t_2)$, $dict.(3,t_2) =1$ at step $11$ and $next$ is enqueued to be
$\langle (3,t_2) \rangle$ at step $12$. After that $frontier$ is equal to $next$. Since
$frontier$ is not empty, we go through step $7$-$12$ again. This time
the forward-neighbor of $(3,t_2)$ is $(3,t_3)$, so $dict.(3,t_3)=2$ and
$next = \langle (3,t_3) \rangle$ and so on. The loop terminates when $frontier$ is empty.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{BFS($G_n, (s,t)$)}{
  $dict.(s,t) = 0$ \\
  $i = 1$ \\
  $fronter = \emptyset$ \\
  \enqueue{$frontier, (s,t)$} \\
 \While{$frontier \ne \emptyset$}{
   $next = \emptyset$ \\
   \For{$(u,t') \in frontier$}{
     \For{$(v,t'') \in$\nei{$(u,t')$}}{
          \If{$(v,t'') \notin d$}{
              $dict.(v,t'') = i$\\
              \enqueue{$next, (v,t'')$}
              }
         }
       }
  $frontier = next$ \\
  $i = i+1$ \\
  }
  \textbf{return} $dict$
}
 \caption{Evolving-graph BFS.
Let $G_n$ be an
evolving graph represented by adjacency lists and $(s, t)$ is a node
of $G_n$.
For each node-time stamp pair $(v_i,t_i)$,
we use a dictionary $dict.(v_i,t_i)$ to store its distance (the length of temporal
paths) from $(s, t)$. We also use
two first-in, first-out queue $frontier$ and $next$, where $frontier$
stores the frontier nodes of the current discovered regions and $next$ stores
the frontier nodes of the next level.
The algorithm returns $dict$ which stores all reachable node-time stamp pairs from $(s,t)$ such that if $dict.(v_i, t_i) = k$,
the length of the shortest temporal path
from $(s,t)$ to $(v_i, t_i)$ is $k$.}
 \label{alg:bfs}
\end{algorithm}

>>>>

\begin{defn}
An \textbf{evolving graph} $G_n$ is a sequence of
(static) graphs
$G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]} \rangle$ with associated
time labels $t_1, t_2, \ldots, t_n$ respectively.
Each $G^{[t]} = (V^{[t]}, E^{[t]})$ represents a (static) graph labeled by a
time $t$.
\end{defn}

@@

We assume no particular relation between the node and edge sets for each static
graph $G^{[t]} = (V^{[t]}, E^{[t]})$. In particular, we allow the node sets to
change over time, so that each $V^{[t]}$ may be different. Changing
node sets happen naturally in citation networks, where nodes may appear or
disappear from the citation network over time.
The addition, removal, or relabeling of nodes can be expressed in terms
of a map $\Pi^{[t,t^\prime]} : V^{[t]} \rightarrow V^{[t^\prime]}$ that expresses
the appropriate permutations and/or projections.

\begin{defn}
A \textbf{temporal node} is a pair $(v, t)$, where $v\in V^{[t]}$ is a node at
a time $t$.
\end{defn}

\begin{defn}
An \textbf{active node} is a temporal node $(v,t)$ if there exists at least one
edge $e\in E^{[t]}$ that connects $v\in V^{[t]}$ to another node $w\in V^{[t]}$,
$w\ne v$.

An \textbf{inactive node} is a temporal node that is not an active node.
\end{defn}

@@

\begin{defn}
A \textbf{temporal path} of length $m$ on an evolving graph $G_n$
from temporal node $(v_1, t_1)$ to temporal node $(v_m, t_m)$} is
a time-ordered sequence of active nodes,
$\langle (v_1, t_1), (v_2, t_2), \ldots, (v_m, t_m) \rangle$.
Here, time ordering means that $t_1 \leq t_2 \leq \cdots \leq t_m$ and
$v_i = v_j$ iff $t_i \ne t_j$.
\end{defn}

This definition of a temporal path differs from that of the dynamic walk
in \cite{gphe11,grihig13} in its explicit requirement of containing only active
nodes. The definition implies that if either or both end points of a temporal
path are inactive, then the entire temporal path must be the empty sequence
$\langle \rangle$. Keeping track explicitly of the time labels of each temporal
node allows greater generality to cases where the node sets change over time.
Furthermore, we shall show later in
%XXX X-REF
that the explicit bookkeeping of the time labels is essential for correctly
generalizing the BFS to evolving graphs.

The following definition of \textbf{forward neighbors} generalizes the notion of
neighbors and reachability in static graphs.

\begin{defn}
The \textbf{$k$-forward neighbors} of a temporal node $(v, t)$ are the temporal nodes
that are the $k+1$st temporal node in every temporal path of length $k+1$ starting
from $(v, t)$. The \textbf{forward neighbors} of a temporal node $(v, t)$ are its
 \textbf{1-forward neighbors}.
\end{defn}

In Figure~\ref{fig:eg_shortest_path},
the forward-neighbors of $(1, t_1)$ are $(2, t_1)$ and $(1, t_2)$ and
the forward-neighbors of $(2, t_1)$ is $(2, t_3)$.
The 2-forward neighbors of $(1, t_1)$ are
$(2, t_1), (1, t_2), (2, t_2)$ and $(3, t_2)$.
By construction, time stamp of every forward neighbor of an active node $(v, t)$
must be no earlier than $t$.

\begin{defn}
The distance from a temporal node $(v, t)$ to a temporal node $(w, s)$ is
the $k$ for which $(w, s)$ is a $k$-forward neighbor of $(v, t)$.
\end{defn}

Note that this notion of distance is not a metric, since the distance from $(v, t)$
to $(w, s)$ will in general differ from the distance of $(v, t)$ from $(w, s)$.

\begin{defn}
A temporal node $(w, s)$ is \textbf{reachable} from a temporal node $(v, t)$ if
the distance to $(w, s)$ from a temporal node $(v, t)$ there exists some finite
integer $k$ for which $(w, s)$ is a $k$-forward neighbor of $(v, t)$.
\end{defn}

The BFS on evolving graphs is described in Algorithm \ref{alg:bfs}.
Given an evolving graph $G_n$ and a root $(s,t)$,
Algorithm~\ref{alg:bfs} returns all temporal notes reachable from the root and
their distances from the root.
The BFS constructs a tree inductively by discovering all $k$-forward neighbors
of the root before proceeding to all $(k+1)$-forward neighbors of the root.

For Figure \ref{fig:eg_shortest_path}, suppose we start with $(1, t_2)$, then
$frontier$ is $\langle(1,t_2)\rangle$ after step $5$.  Since the only forward-neighbor of
$(1,t_2)$ is $(3,t_2)$, $dict.(3,t_2) =1$ at step $11$ and $next$ is enqueued to be
$\langle (3,t_2) \rangle$ at step $12$. After that $frontier$ is equal to $next$. Since
$frontier$ is not empty, we go through step $7$-$12$ again. This time
the forward-neighbor of $(3,t_2)$ is $(3,t_3)$, so $dict.(3,t_3)=2$ and
$next = \langle (3,t_3) \rangle$ and so on. The loop terminates when $frontier$ is empty.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{BFS($G_n, (s,t)$)}{
  $dict.(s,t) = 0$ \\
  $i = 1$ \\
  $fronter = \emptyset$ \\
  \enqueue{$frontier, (s,t)$} \\
 \While{$frontier \ne \emptyset$}{
   $next = \emptyset$ \\
   \For{$(u,t') \in frontier$}{
     \For{$(v,t'') \in$\nei{$(u,t')$}}{
          \If{$(v,t'') \notin d$}{
              $dict.(v,t'') = i$\\
              \enqueue{$next, (v,t'')$}
              }
         }
       }
  $frontier = next$ \\
  $i = i+1$ \\
  }
  \textbf{return} $dict$
}
 \caption{Breadth-first search (BFS) on an evolving graph: given
$G_n$, an evolving graph represented by adjacency lists, and a root $(s, t)$,

 is a node
of $G_n$.
For each node-time stamp pair $(v_i,t_i)$,
we use a dictionary $dict.(v_i,t_i)$ to store its distance (the length of temporal
paths) from $(s, t)$. We also use
two first-in, first-out queue $frontier$ and $next$, where $frontier$
stores the frontier nodes of the current discovered regions and $next$ stores
the frontier nodes of the next level.
The algorithm returns $dict$ which stores all reachable node-time stamp pairs from $(s,t)$ such that if $dict.(v_i, t_i) = k$,
the length of the shortest temporal path
from $(s,t)$ to $(v_i, t_i)$ is $k$.}
 \label{alg:bfs}
\end{algorithm}
